\myTitle

### 2025-03-24 tex/responses-to-referees/responses-to-reviewers-1.Rmd

* In addition, L239-240 implies you were you only interested in right side of the peak. I would assume you were interested in up and downstream of the peak unless I missed something. In addition, please clarify why 2.2 Mb selected. Is this an LD block? Does this mean 2.2 Mb in each direction or total?
    - _It turns out that the main peak of association is at the
    beginning of chromosome 34, so there is no flanking region to the left of the peak. To
    clarify why those candidate SNPs in the flanking region were chosen we now state:_ 
    "The final condition was implemented in order to gather several SNPs with large allele frequency differences
    that were adjacent to, but not directly within, the main peak of association, so as to possibly learn about
    recombination in the region."

* The writing of the manuscript is largely well written but makes high use of words such as “we” or “us”, which although stylistic, makes the manuscript longer than needed in some cases such as “we then dropped” or “this allows us”. In many cases these can be removed (examples in minor comments).
    - _We addressed all the examples in the minor comments and also removed additional instances of the
    use of such language where it was practical._

* There is also use of informal language e.g. “amplified so well” or unnecessary words such as “then” that could be removed.
    - _Following this suggestion, we removed many occurrences of "then" and "as well as"_

* I commend the use of the distribution of log-likelihood ratio for relationship to show the power of haplotypes vs SNPs and would argue this could be elevated to a full figure, potentially in a condensed form, maybe less panels somehow?
    - _We appreciate the referee's feedback on this.  In response, we have created a
    new figure in the paper itself that examines the distribution of log-likelihood
    ratios only in the Feather River Hatchery.  This figure distinguishes between
    different log-likelihood ratios (PO/U, FS/U, and HS/U) and nicely shows the
    advantages of using microhaplotypes for relationship inference. We have also left
    the original figure in the supplement for any readers curious about the effect
    in different collections._

* If possible, I think it would also be good to show how much of an improvement the discrimination power improved from the old marker data. In theory, this data has already been generated for the full population if you strip out the novel SNPs of the current study.
    - _After long and careful consideration of this, we determined that it would not
    be possible to make an accurate comparison between the old and the new marker
    data in this regard.  Unfortunately, only 78 of the original 94 TaqMan assays
    were able to be converted into amplicon-sequenced markers, so we are not able
    to reproduce the old marker set with the same reference samples._

@@

- **Figure 1 and 2: The colours for “spring run” and “Late-fall run” are very difficult to separate for a colour-blind person. Since the groups are not in a specific order on the graph, it is very hard work to out what is what. I’d recommend the authors use a selection of colours that will work for colour-blind people.**

    - **Planning to just cross-hatch the light blue of the late fall**

- Line 589: “unprecedented” is an over-statement as there is no independent validation set. The authors should use a word that less forceful.
    - _Changed to "excellent"_

- Sentence at lines 645-647. Can the authors provide a reference to back up this statement.
    - _We added a citation here that extensively describes the use of a Fluidigm chip-based panel for
    parent pair allocation in Chinook salmon in California._

### 2025-03-24 003-ckmr-sim-evaluation.Rmd
Getting close to finishing reponses to referees

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/SWFSC-chinook-reference-baseline.csv.gz",
    pop_labels = "inputs/reference-collection-names.csv",
    locus_info = "inputs/Calif-Chinook-Amplicon-Panel-Information.csv"
  )
  output_list <- list(
    ckmr_figure = "results/ckmr-sim-eval/fpr-fnr-figure.pdf",
    ckmr_figure_crop = "results/ckmr-sim-eval/fpr-fnr-figure-crop.pdf",
    ckmr_figure_tex = "tex/images/fpr-fnr-figure-crop.pdf",
    ckmr_comp = "results/ckmr-sim-eval/cmkr-comp-figure.pdf",
    ckmr_comp_crop = "results/ckmr-sim-eval/cmkr-comp-figure-crop.pdf",
    ckmr_comp_tex = "tex/images/cmkr-comp-figure-crop.pdf"

@@

# now, we also want to get the genome positions of things.
# we take its position to be the start position of the amplicon
gen_coords <- read_csv(input_list$locus_info) %>%
  rename(Locus = AmpliconName, Chrom = Otsh_v2.0_chromosome, Pos = AmpliconStartPos) %>%
  select(Locus, Chrom, Pos)

@@

```{r}
#' @param X a tibble like that nested in a single row of ckmr_sim_nests
#' @param GC2 essentially the gen_coords object from above
ckmr_afreqs_ready <- function(X, GC2) {
  tmp <- X %>%
    select(-(repunit:sample_type)) %>%
    pivot_longer(
      -indiv,
      names_to = c("Locus", "gene_copy"),
      values_to = "Allele",
      names_pattern = "^(.*)_([12])$"
    ) %>%
    filter(!is.na(Allele)) %>%
    count(Locus, Allele) %>%
    group_by(Locus) %>%
    filter(n() > 1) %>% # filter out monomorphic loci
    mutate(
      Freq = n / sum(n),
    ) %>%
    ungroup() %>%
    left_join(GC2, by = join_by(Locus)) %>%
    select(Chrom, Pos, Locus, Allele, Freq) %>%
    arrange(Pos, desc(Freq))

@@

source("R/shift_legend.R")

@@

a <- grid.arrange(shift_legend(g))
```

@@

a2 <- grid.arrange(shift_legend(g2))

@@

Neil suggested, and Carlos and I agreed, that one way to see they improvement in
relationship inference obtained by using multiallelic microhaps is to look at the
separation of the different relationship log-likelihood ratios when using just the
best (MAF closest to 0.5) SNP from each locus, or when using all the SNPs/alleles.
We deciced to do this for the hatchery populations in California, as well as the
winter run.  I am going to implement this in a few functions, each of which operate
on the one of the ckmr_sim_nests.

@@

  # now extract the logl_ratios for PO vs U and return those
  extract_logls(Q_link, numer = c(PO = 1), denom = c(U = 1))
}
```

@@

```{r}
# collections / groups we want to do
pops <- c("ck-FRH", "ck-IGH", "ck-SRW", "ck-TRH")

@@

>>>>

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/SWFSC-chinook-reference-baseline.csv.gz",
    pop_labels = "inputs/reference-collection-names.csv",
    locus_info = "inputs/Calif-Chinook-Amplicon-Panel-Information.csv"
  )
  output_list <- list(
    ckmr_figure = "results/ckmr-sim-eval/fpr-fnr-figure.pdf",
    ckmr_figure_crop = "results/ckmr-sim-eval/fpr-fnr-figure-crop.pdf",
    ckmr_figure_tex = "tex/images/fpr-fnr-figure-crop.pdf",
    ckmr_comp = "results/ckmr-sim-eval/cmkr-comp-figure.pdf",
    ckmr_comp_crop = "results/ckmr-sim-eval/cmkr-comp-figure-crop.pdf",
    ckmr_comp_tex = "tex/images/cmkr-comp-figure-crop.pdf",
    frh_comp = "results/ckmr-sim-eval/frh-comp-figure.pdf",
    frh_comp_crop = "results/ckmr-sim-eval/frh-comp-figure-crop.pdf",
    frh_comp_tex = "tex/images/frh-comp-figure-crop.pdf"

@@

# now, we also want to get the genome positions of things.
# we take its position to be the start position of the amplicon
gen_coords <- read_csv(input_list$locus_info) %>%
  rename(Locus = AmpliconName, Chrom = Otsh_v2.0_chromosome, Pos = AmpliconStartPos) %>%
  select(Locus, Chrom, Pos) %>%
  filter(!str_detect(Locus, "sexy"))

@@

```{r}
#' @param X a tibble like that nested in a single row of ckmr_sim_nests
#' @param GC2 essentially the gen_coords object from above
ckmr_afreqs_ready <- function(X, GC2) {
  tmp <- X %>%
    select(-(repunit:sample_type)) %>%
    pivot_longer(
      -indiv,
      names_to = c("Locus", "gene_copy"),
      values_to = "Allele",
      names_pattern = "^(.*)_([12])$" 
    ) %>%
    filter(!is.na(Allele)) %>%
    count(Locus, Allele) %>%
    group_by(Locus) %>%
    filter(n() > 1) %>% # filter out monomorphic loci
    mutate(
      Freq = n / sum(n),
    ) %>%
    ungroup() %>%
    filter(!is.na(Locus)) %>%  # this gets rid of the rosa stuff
    left_join(GC2, by = join_by(Locus)) %>%
    select(Chrom, Pos, Locus, Allele, Freq) %>%
    arrange(Pos, desc(Freq))

@@

@@

library(lemon) 
a <- reposition_legend(g, panel = "panel-2-3",  x = 0, y = 0, just = c(0,0))
```

@@

a2 <- reposition_legend(g2, panel = "panel-2-3", x = 0, y = 0, just = c(0,0))
```

@@

Neil suggested, and Carlos and I agreed, that one way to see the improvement in
relationship inference obtained by using multiallelic microhaps is to look at the
separation of the different relationship log-likelihood ratios when using just the
best (MAF closest to 0.5) SNP from each locus, or when using all the SNPs/alleles.
We deciced to do this for the hatchery populations in California, as well as the
winter run.  I am going to implement this in a few functions, each of which operate
on the one of the ckmr_sim_nests.

@@

  # now extract the logl_ratios
  extract_logls(Q_link, numer = c(PO = 1), denom = c(U = 1))
}
```

@@

```{r}
# collections / groups we want to do.  I originally did many, but
# just whittled it down to include in main text following a reviewer request.
pops <- c("ck-FRH", "ck-IGH", "ck-SRW", "ck-TRH")

@@

## Making a comparison figure for the main text

Reviewer one wanted to see this in the main text.  I will show just a single
figure for FRH.  I also want to look at PO, FS, and FS versus U specifically.
```{r}
pop <- c("ck-FRH")

# let's put the microhaps and the best SNPs together in a long format.
input_tibbles2 <- bind_rows(
  ckmr_sim_nests %>%
    filter(ckmr_sim_group %in% pop) %>%
    mutate(condition = "microhaps", .before = ckmr_sim_group),
  ckmr_sim_nests %>%
    filter(ckmr_sim_group %in% pop) %>%
    mutate(tops = map(data, retain_top_snp)) %>%
    mutate(data = tops) %>%
    select(-tops) %>%
    mutate(condition = "top_snps", .before = ckmr_sim_group)
)

# a new function that gets multiple logls out of it
get_logl_rats_multi <- function(X, GC2 = gen_coords) {

  # now extract the logl_ratios for PO, FS, or HS vs U and return those
  list(
    POU = extract_logls(Q_link, numer = c(PO = 1), denom = c(U = 1)),
    FSU = extract_logls(Q_link, numer = c(FS = 1), denom = c(U = 1)),
    HSU = extract_logls(Q_link, numer = c(HS = 1), denom = c(U = 1))
  ) %>%
    bind_rows(.id = "ratio_type")
}

Qsims2 <- lapply(input_tibbles2$data, get_logl_rats_multi)

AllSims2 <- tibble(
  condition = input_tibbles2$condition,
  ckmr_sim_group = input_tibbles2$ckmr_sim_group,
  Qsims = Qsims2
) %>% 
  unnest(Qsims) %>%
  filter(  # only keep the relevant ones
    (ratio_type == "POU" & true_relat %in% c("U", "PO")) |
      (ratio_type == "FSU" & true_relat %in% c("U", "FS")) |
      (ratio_type == "HSU" & true_relat %in% c("U", "HS"))
  )
```

Now, we make a picture of that.
```{r}
library(ggpattern)
AS3 <- AllSims2 %>%
  mutate(
    ratio_type = case_match(
      ratio_type,
      "POU" ~ "Parent-offspring versus Unrelated log-likelihood ratio",
      "FSU" ~ "Full-sibling versus Unrelated log-likelihood ratio",
      "HSU" ~ "Half-sibling versus Unrelated log-likelihood ratio"
    ),
    ratio_type = factor(
      ratio_type,
      levels = c(
        "Parent-offspring versus Unrelated log-likelihood ratio",
        "Full-sibling versus Unrelated log-likelihood ratio",
        "Half-sibling versus Unrelated log-likelihood ratio"
      )
    )
  ) %>%
  rename(`True\nRelationship` = true_relat)

g8 <- ggplot(AS3, aes(x = logl_ratio, fill = `True\nRelationship`, alpha = condition)) +
  geom_density() +
  scale_alpha_manual(values = c(microhaps = 0.7, top_snps = 0.3)) +
  facet_wrap(~ratio_type, ncol = 1, scales = "free") +
  xlab("Log-likelihood Ratio") +
  ylab("Density") +
  theme_bw() +
  theme(legend.position = "top") +
  guides(alpha = "none")

ggsave(g8, filename = output_list$frh_comp, width = 4, height = 6)

# let's crop that down for use, too
CALL <- paste("pdfcrop ", output_list$frh_comp, collapse = " ")
system(CALL)

file.rename(from = output_list$frh_comp_crop, to = output_list$frh_comp_tex)

### 2025-03-24 tex/body-text-of-paper.tex
Getting close to finishing reponses to referees

Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.
Gene regions were selected that (1) contained multiple SNPs within a 100 bp window, and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.

These gene regions were PCR-amplified and sequenced on a MiSeq sequencer using a $2\times 75$-cycle kit.
The sequences from each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignmets were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

The presence of multiple alleles at many of these loci improves power for
kin inference compared to a marker panel using only a single SNP at each
amplicon.  To provide a visual display of the difference afforded by using
microhaplotypes compared to single SNPs in this data set, we show the
distribution of log-likelihood ratios for different relationships using our microhaplotype-scored
amplicons, versus using the most heterozygous single SNP from each
amplicon (\ref{fig:ckmr-comp}).  The figures show a considerable increase
in separation between the relationships categories from calling genotypes
as microhaplotypes in these amplicons.

@@

Previous genetic marker sets for California Chinook salmon had sufficient power for inferring
parent-offspring relationships, but only when both parents were sampled and genotyped.
This marker set considerably increases the capacity for relationship inference in California
salmon, by providing sufficient power for parent-offspring relationship inference, when only
one parent is sampled, and identifying pairs of full siblings when no parents are sampled.
Groups of full siblings larger than two can be identified even more accurately than expressed in Figure~\ref{fig:fprs},
when using statistical approaches that account for the joint relationships between more than two individuals,
such as COLONY \citep{wang2004sibship}.
Although some additional errors might be
expected when related but non-target kin pairs are sampled,  the false positive rates
for these non-target kin pairs are sufficiently small
for almost all realistic scenarios.
Accurate relationship
inference in even the most genetically depauperate Chinook salmon population in California
and rangewide \citep{seeb2007development,clemento2014evaluation} is therefore possible with this marker set.
We note that, as in previous work, it is nearly impossible to identify HS pairs accurately
with data from this, or any standard marker dataset. Accurate identification of HS pairs
typically relies on many hundreds of microhaplotype markers \citep{baetscher2018microhaplotypes} or thousands of SNP markers \citep{hillary2018genetic}.

>>>>

Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were then tested for PCR-amplification and consistency.  This step included
amplification in thousands of samples from a number of different collections of Chinook salmon for sampled
for several projects, with
subsequent sequencing on a
MiSeq sequencer using a $2\times 75$-cycle kit.
The sequences from each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

The presence of multiple alleles at many of these loci improves power for
kin inference compared to a marker panel using only a single SNP at each
amplicon.  To provide a visual display of the difference afforded by using
microhaplotypes compared to single SNPs in this data set, we show the
distribution of log-likelihood ratios for different relationships using our microhaplotype-scored
amplicons, versus using the most heterozygous single SNP from each
amplicon for all fish from the FRH (Fig.~\ref{fig:ckmr-comp}), with results for the PO vs Unrelated
log likelihood ratio shown for more collections in Fig.~\ref{fig:ckmr-comp-supp}.  The figures show a considerable increase
in separation between the relationship categories from calling genotypes
as microhaplotypes in these amplicons.
%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\newcommand{\fprcap}{\footnotesize The distribution of log-likelihood ratios for
fish from Feather River Hatchery when scoring markers as microhaplotypes versus as
single SNPs.  Three different likelihood ratios are shown in the three panels.  The opacity
of the distributions indicates the marker type: the more opaque setting indicates markers
scored as microhaplotypes; the more transparent setting indicates the use of only the most
heterozygous SNP from amongst all SNPs in the microhaplotype.  Evidently, scoring the markers
as microhaplotypes produces more separation between the distributions for the relationships of PO, FS, or HS, against
the relationship of unrelated (U).}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/frh-comp-figure-crop.pdf}
\end{center}
\caption[\fprcap]{\fprcap}
\label{fig:ckmr-comp}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

@@

Previous genetic marker sets for California Chinook salmon had sufficient power for inferring
parent-offspring relationships, but only when both parents were sampled and genotyped \citep{clemento2013creation}.
This marker set considerably increases the capacity for relationship inference in California
salmon, by providing sufficient power for parent-offspring relationship inference, when only
one parent is sampled, and identifying pairs of full siblings when no parents are sampled.
Groups of full siblings larger than two can be identified even more accurately than expressed in Figure~\ref{fig:fprs},
when using statistical approaches that account for the joint relationships between more than two individuals,
such as COLONY \citep{wang2004sibship}.
Although some additional errors might be
expected when related but non-target kin pairs are sampled,  the false positive rates
for these non-target kin pairs are sufficiently small
for almost all realistic scenarios.
Accurate relationship
inference in even the most genetically depauperate Chinook salmon population in California
and rangewide \citep{seeb2007development,clemento2014evaluation} is therefore possible with this marker set.
We note that, as in previous work, it is nearly impossible to identify HS pairs accurately
with data from this, or any standard marker dataset. Accurate identification of HS pairs
typically relies on many hundreds of microhaplotype markers \citep{baetscher2018microhaplotypes} or thousands of SNP markers \citep{hillary2018genetic}.
